---
title: "main"
format: html
editor: visual
---

# Introduction

This R project documents analysis of Ephemeroptera, Plecoptera, and Trichoptera larva collected in relation to a field study conducted in Trøndelag, Norway. Benthic invertebrate sampling was performed in 6 lakes, 3 regulated for hydropower production and 3 controls, every 2nd week during spring and summer of 2023 and 2024. A standardized kick sampling method was used, with 1-minute intervals. Sampling started as soon after ice melt as possible and ended in late June/early July, resulting in 4 sampling times. See the document README.md for further details on licenses and citation.

Corresponding publication:

> **Paper Title:** *Hydropower regulation reshapes littoral invertebrate communities: functional traits reveal impacts beyond richness and abundance*\
> **Authors:** *Krogstie, Hanne Bjørnås; Solvåg, Christianne Dalsbotten; Ratikainen, Irja Ida; Kjærstad, Gaute; Finstad, Anders Gravbrøt*\
> **Journal:** *Accepted to Aquatic Sciences* (2026)\
> **DOI:** *10.1007/s00027-026-01367-3*

# 1. Setup

Run setup script, which loads (and installs if needed) packages.

```{r load-packages, warning=FALSE}
# Load packages
source("1_setup.R")
```

# 2. Load data

The scripts 2_1 to 2_7 loads, structures, and tidies raw data, which then can be used for analysis or data visualization.

**Considering sample variability and removing replicates**

Multiple kick sampling replicates (5 per event in 2023 and 3 per event in 2024) was taken per event, with the initial goal of processing, sorting and identifying all. Early in the labwork period, multiple replicates were therefore sorted as a part of the process of assessing available time for labwork, processing time, and sample variability within a event. Based on an investigation of sample variability within the first sampling event (see the script `6_1_sample_variability_check.R` ) and the available time and resources, only one sample per event was consistently processed for the remaining material.

In script `2_7_data_tidy_and_merge.R` we therefore remove two additional replicates for the sampling event on the date 24.05.2023 from lake Stor-Drakstsjøen, which were processed prior to this decision, to ensure consistent sampling effort across all lakes in our analyses.

```{r load-data}
# 2_1
source("2_1_data_occurrences.R")
# 2_2
source("2_2_data_trait_locomotion.R")
# 2_3
source("2_3_data_trait_armour.R")
# 2_4
source("2_4_data_temperature.R")
# 2_5
source("2_5_data_water_levels.R")
# 2_6
source("2_6_data_hydropower.R")
# 2_7
source("2_7_data_tidy_and_merge.R")
```

# 3. Analyses

The scripts 3_1 to 3_7 contain the code necessary to run the analyses performed in the publication. Each script contains further details.

```{r analyses}
# 3_1
source("3_1_species_abundance.R")
# 3_2
source("3_2_species_richness.R")
# 3_3
source("3_3_community_composition.R")
# 3_4
source("3_4_betadiversity_partitioning.R")
# 3_5
source("3_5_indicator_species.R")
# 3_6
source("3_6_trait_locomotion.R")
# 3_7
source("3_7_trait_armour.R")
```

# 4. Figures and tables

The scripts 4_1 to 4_5 can be run to create the majority of figures and tables included in the publication. The naming of each script reflects the figure or table number in the publication.

```{r figures-tables}
# 4_1
source("4_1_trait_armour.R")
```

# 5. Supplementary figures and tables

The scripts 5_1 to 5_8 can be run to create the majority of figures and tables included in the Supplementary Information (SI) of the publication. The naming of each script reflects the figure or table number in the SI of the publication.

```{r SI-figures-tables}

```

# 6. Additional analyses

The scripts `6_1_sample_variability_check.R` and `6_2_spatial_distance_check.R` documents additional analyses done in parallel with the general analysis workflow.

```{r additional-analyses}

```
